Swinging, Fast and Slow:\\Interpreting variation in baseball swing tracking metrics

### 2025-06-26 reports/arxiv/paper.tex
SP comments and edits

    We reproduce MLB's derived metric for {\it squared-up} contact as:
    \begin{equation}
        \label{eqn:squared-up}
    \mbox{batted ball speed} > 80\% \times \left\{1.23 \cdot (\mbox{bat speed}) + 0.23 \cdot (\mbox{pitch speed})\right\}.
    \end{equation}
    The right-hand side of Equation \ref{eqn:squared-up} is a rough approximation for 80\% of the theoretical maximum batted ball speed given bat speed and pitch speed at contact, with additional assumptions about the mass of bat and ball \citep{nathan_dynamics_2000}. In our dataset, we observe batted ball speeds as high as 110\% of the rough theoretical maximum.

@@

    To this end, we fit the following skew-normal (SK) multilevel model:

    \begin{align}
    \label{eqn:intention-swing-length}
    \begin{split}
        ( \mbox{swing length} )_i &\sim \mbox{SK}(\mu_i, \sigma, \alpha_i) \\
        \mu_i &= \mu_0 + \gamma_{p_i} + \gamma_{b_i}
        + \beta^B \cdot (\mbox{balls})_i
          + (\beta^S + \gamma^S_{b_i}) \cdot (\mbox{strikes})_i\\
        & + (\beta^X + \gamma^X_{b_i}) \cdot (\mbox{pitch loc x})_i
          + (\beta^Z + \gamma^Z_{b_i}) \cdot (\mbox{pitch loc z})_i \\
          \alpha_i &= \alpha_0 + \nu_{b_i}
    \end{split}
    \end{align}
    where $b_i$ denotes the batter on swing $i$ against pitcher $p_i$. We use the same specification for both swing length and bat speed intention models.

    In detail, the $\gamma$ parameters are random intercepts and random slopes describing the mean of the swing distribution such that,

    \begin{align}
        \begin{split}
            \gamma_{p_i} &\sim \mathcal{N}(0, \sigma^2_p) \\
            \boldsymbol{\gamma}_{b_i} &\sim \mathcal{N}(\boldsymbol{0}, \Sigma)
        \end{split}
    \end{align}
    where $\boldsymbol{\gamma}_{b_i}$ is the vector of batter random intercept and slopes $(\gamma_{b_i}, \gamma^S_{b_i}, \gamma^X_{b_i}, \gamma^Z_{b_i})$ which follow a multivariate normal distribution centered around the zero vector $\boldsymbol{0}$ with covariance matrix $\Sigma$. This $\Sigma$ contains the relevant random effect variances for the batter-level terms, $(\sigma_b, \sigma_b^S, \sigma_b^X, \sigma_b^Z)$, as well as their respective correlations. This allows us to quantify variation between hitters and pitchers at the intercept level, as well as capture variation in batter approach and swing adaptation.

@@

    We fit both bat speed and swing length models in a Bayesian framework via the \texttt{brms} package in \texttt{R} \citep{brms}, which provides an interface for Bayesian modeling with Stan \citep{carpenter2017stan}. We use weakly informative priors for the parameters, with vague half-$t_3$ priors (i.e., a Student's $t$ distribution centered at zero with 3 degrees of freedom but truncated to positive values) for the standard deviation parameters \citep{gelman2006prior} and the LKJ prior for the correlations between batter-level effects \cite{lewandowski2009generating}. Our Bayesian approach naturally provides uncertainty quantification for the model parameters via their posterior distributions, which are estimated using MCMC sampling. For model fitting, we use four parallel chains with 6,000 (3,000 burn-in) and 4,000 (2,000 burn-in) samples for the swing length and bat speed models respectively. We observe evidence of convergence of the MCMC algorithm based on observing trace plots and $\hat{R}$ statistics close to 1 \citep{gelman1992inference, brooks1998general}, along with no indication of problematic effective sample sizes \citep{gelman2020bayesian}.

    For comparison, we also fit simpler Gaussian versions of the swing length and bat speed models. These models follow a specification similar to that of the SK models in Equation \ref{eqn:intention-swing-length}, but with the shape parameter removed (i.e., the Guassian distribution is a SK with shape equal to 0). We use an 80/20 train/test split to compare the out-of-sample performance of the SK and Gaussian fits against each other. We compute the expected log pointwise predictive density (ELPD) on the 20\% test data, which involves computing the log-likelihood of each test observation for each posterior sample. In addition to providing us with standard errors about the difference in model performance, we choose ELPD as our evaluation metric because of our interest in comparing the skewness of the SK posterior predictive distributions relative to the symmetric Gaussian.

@@

    The out-of-sample ELPD results are shown in Table \ref{tab:intention-model-elpd} for both swing length and bat speed models. For both settings, the SK model displayed an ELPD of around 3 standard errors better than the Gaussian model. The remaining results in this manuscript are based on the SK intention models.

\begin{table}[H]
        \centering
        \begin{tabular}{l|r|r|r|}
          & $\Delta$ELPD  & SE & \# SEs \\
          \hline
          \input{../../tables/intention_elpd.tex}
        \end{tabular}
        \caption{\it Model comparison via expected log predictive density (ELPD). Shown are the difference in EPLD between the skew-normal (SK) and Gaussian models ($\Delta$ELPD), the standard error of ELPD difference (SE), and the number of standards errors the SK model is better than the Gaussian model.}
        \label{tab:intention-model-elpd}
      \end{table}

       Table \ref{tab:fixed-effects} summarizes the fixed effects of the bat speed and swing length intention models. We observe intuitive relationships for both models with regards to the relationships between pitch-level context and each observed swing measurement. As the number of balls increase, on average ($\beta^B$) we observe faster and longer swings. The opposite behavior is observed as the number of strikes increases ($\beta^S$), with slower and shorter swings on average. Figure \ref{fig:approach} displays the joint distribution of the posterior mean for the batter random slopes for strikes ($\hat{B}^S + \hat{\gamma}_b^S$) in both intention models, regardless of the individual batter effect we expect a decrease in bat speed and swing length per strike. A similar relationship is observed for pitches with higher vertical locations ($\beta^Z$). The two models diverge with regards to the horizontal location of pitches ($\beta^X$), as pitches that thrown further away from the batter result in longer but slower swings on average. With regards to the shape parameter of the SK distribution, the intercepts ($\alpha_0$) for both bat speed and swing length are negative which indicate left skewed distributions.

      \begin{figure}[H]
        \centering
        \includegraphics[width = 0.6\textwidth]{../../figures/approach.pdf}
        \caption{\it Estimated batter random slopes for strikes in the intention model. Each point is a batter $b$, with the x-value representing $\hat\beta^S + \hat\gamma^S_b$ from the model (\ref{eqn:intention-swing-length}) and the y-value representing the same quantity from the model (\ref{eqn:intention-bat-speed}). This quantity is interpretable as the expected change in swing length (or bat speed) corresponding to a one-strike increase in the ball-strike count.}
        \label{fig:approach}
      \end{figure}

@@

      Next, Table \ref{tab:intention-variances} summarizes the posterior distributions of the random effects variance terms. For both bat speed and swing length intention models, we observe the highest variation is attributed to differences between batters ($\sigma_b$). Interestingly, the variation between batters in the shape random effects ($\tau_b$) is the second highest for swing length but is smaller than the pitch location random slopes ($\sigma_b^X$ and $\sigma_b^Z$) for the bat speed model. For each of the estimated terms, all of the 95\% credible intervals are strictly greater than zero demonstrating evidence of variability between batters at the intercept and slope-level.

@@

      Figure \ref{fig:intent-re} provides an overview of the batter-level $\mu_i$ random effects in the SK intention models. The variation we observe between batters aligns with intuition, as batters known for taking longer/faster swings (e.g., Oneil Cruz, Giancarlo Stanton, Javier Baez) display higher random intercepts ($\gamma_b$) in compare to batters with shorter/slower swings (e.g., Luis Arraez). We observe positive correlations between the random slopes of bat speed and swing length, providing information on the variation between batters with respect to their approach ($\gamma_b^S$) and adaptation ($\gamma_b^X$, $\gamma_b^Z$). For example, Anthony Rizzo's approach appears to be more conservative with the biggest decrease in bat speed and swing length for each additional strike relative to the average batter slope. We also observe batters that deviate from the consistent pattern between bat speed and swing length, such as Shohei Ohtani with regards to his horizontal pitch location random slope appearing to be higher than average for swing length despite a lower than average value for bat speed.

      \begin{figure}[H]
        \centering
        \includegraphics[width = 0.8\textwidth]{../../figures/intent_re.pdf}
        \caption{\it Joint distribution of the posterior means for the batter-level $\mu_i$ random effects in the SK intention models for both bat speed and swing length. Each batter is displayed by their name, and only batters with at least 25 squared-up swings are displayed.}
        \label{fig:intent-re}
      \end{figure}

       In contrast, Figure \ref{fig:alpha-re} displays the joint distribution of the posterior means for the batter random intercepts in modeling the shape parameter $\alpha$ of the SK distribution. Unlike the mean-level random effects, we do not observe correlation between the bat speed and swing length shape-level random intercepts. The variation captured by these random intercepts provides some understanding of the differences in the intended swing distribution for players. Luis Arraez displays above average random intercepts for both bat speed and swing length, indicating that his intention distributions are less skewed left than compared to other batters.

@@

      The previous section demonstrates that there are differences between batters in how they modulate their swing length and bat speed as the number of strikes in the count increases. This section addresses the consequences of those differences between batters. As described in Section \ref{sec:methods-causal}, we estimate three separate regression models with instrumental variables for contact probability given swing (\ref{eqn:causal-model-contact}), fair ball probability given contact (\ref{eqn:causal-model-fair}) and expected xwOBAcon given fair ball (\ref{eqn:causal-model-hit}). The estimated effects of bat speed appraoch and swing length approach from these models are reported in Table \ref{tab:causal-model}.

      \begin{table}[H]
        \centering
        \begin{tabular}{l|r|r|r|}
          & Contact Model (\ref{eqn:causal-contact}) & Fair/Foul Model (\ref{eqn:causal-fair}) & xwOBAcon Model (\ref{eqn:causal-hit})\\
          \hline
          \input{../../tables/causal_model.tex}
        \end{tabular}
        \caption{\it Estimated regression coefficients and corresponding standard errors from the causal models (\ref{eqn:causal-contact})--(\ref{eqn:causal-hit}). Bat Speed Approach is defined as the batter's change in bat speed per strike added to the count, as estimated by (\ref{eqn:intention-bat-speed}). Swing Length Approach is defined analogously as the batter's change in swing length per strike added to the count, as estimated by (\ref{eqn:intention-swing-length}).}
        \label{tab:causal-model}
      \end{table}

@@

    In this work, we use newly available bat tracking data and examine the relationship between swing aggression and outcomes in baseball. We recognize that both bat speed and swing length are inherently outcomes of the swing, confounded by whether or not the batter identified the pitch. To address this, we first model the batter's bat speed and swing length intent via Bayesian hierarchical skew-normal models with evidence of more appropriate fits compared to simpler Gaussian models. Our results demonstrate the variation between batters in multiple aspects, including how they change their swing metrics as the number of strikes increases. We then condition on the batter's count-intended bat speed and swing length as instrumental variables to estimate the causal effect of these intended swing metrics on swing outcomes. The results are intuitive, indicating that batters who shorten and slow down their swing more per strike are expected to make contact more often in 1-strike and 2-strike counts relative to other batters. However, we observe a trade-off as this type of conservative approach at the plate is associated with less power. We quantify this trade-off in terms of run value, revealing on an interpretable scale that the highest-value approaches cut down on swing length without sacrificing bat speed.

@@

    While Section \ref{sec:results-value} illustrated the tradeoff involved with how batters vary their \textit{adaptation}, there are of course other sources of variation in bat speed and swing length. For example, in Section \ref{sec:methods-intention} we define \textit{adaptation} as the extent that a batter's swing tracking metrics co-vary with pitch location. As captured by $\sigma_b^X$ and $\sigma_b^Z$ in Table \ref{tab:intention-variances} and the posterior mean for the batter level random effects in Figure \ref{fig:intent-re}, we observe differences between batters in their intended \textit{adaptation} as the pitch location varies horizontally and vertically. To demonstrate the variation, Figure \ref{fig:adaptation} displays the expected swing length from the intention model by pitch location for two sample batters, Nico Hoerner and Juan Soto (assuming a 0-0 count). Hoerner displays a clear relationship with the vertical location with shorter intended swings on average for pitches that are located higher in the zone. In comparison, Soto displays a fairly consistent swing length throughout the zone with an uptick in length for low and away pitches, versus a shortening of the swing for high and inside pitches. Understanding why batters display such variation in their \textit{adaptation} is left for future work, and could be explored with swing path measurements describing the tilt and attack angle that were released during the 2025 MLB season \citep{petriello_4_2025}.

@@

      We limited our analysis of the run value of approach to a batter of average skill, but it is possible that the results in Section \ref{sec:results-value} could differ depending on the profile of the batter. For example, the value of a specific approach could be different for high-contact, low-power batters versus low-contact, high-power batters. Future work could estimate personalized effects of batter approach based on the batter's strengths and weaknesses. Additionally, the estimated run value was based on linear weights, which ignore the base-out situation. With a runner on third base on less than two outs, the value of putting the ball in play (relative to non-ball-in-play outcomes) increases, so presumably the trade-off of increased contact for decreased power is more valuable in this situation. Future work could estimate this effect and evaluate the extent to which batters tailor their approach to the base-out situation.

>>>>

    We reproduce MLB's derived metric for {\it squared-up} contact as:
    \begin{equation}
        \label{eqn:squared-up}
    \mbox{batted ball speed} > 80\% \times \left\{1.23 \cdot (\mbox{bat speed}) + 0.23 \cdot (\mbox{pitch speed})\right\}.
    \end{equation}
    The right-hand side of Equation (\ref{eqn:squared-up}) is a rough approximation for 80\% of the theoretical maximum batted ball speed given bat speed and pitch speed at contact, with additional assumptions about the mass of bat and ball \citep{nathan_dynamics_2000}. In our dataset, we observe batted ball speeds as high as 110\% of the rough theoretical maximum.

@@

    To this end, we fit the following skew-normal (SK) multilevel model:
    \begin{align}
    \label{eqn:intention-swing-length}
    \begin{split}
        ( \mbox{swing length} )_i &\sim \mbox{SK}(\mu_i, \sigma, \alpha_i) \\
        \mu_i &= \mu_0 + \gamma_{p_i} + \gamma_{b_i}
        + \beta^B \cdot (\mbox{balls})_i
          + (\beta^S + \gamma^S_{b_i}) \cdot (\mbox{strikes})_i\\
        & + (\beta^X + \gamma^X_{b_i}) \cdot (\mbox{pitch loc x})_i
          + (\beta^Z + \gamma^Z_{b_i}) \cdot (\mbox{pitch loc z})_i \\
          \alpha_i &= \alpha_0 + \nu_{b_i}
    \end{split}
    \end{align}
    where $b_i$ denotes the batter on swing $i$ against pitcher $p_i$. We use the same specification for both swing length and bat speed intention models.

    In detail, the $\gamma$ parameters are random intercepts and random slopes describing the mean of the swing distribution such that
    \begin{align}
        \begin{split}
            \gamma_{p_i} &\sim \mathcal{N}(0, \sigma^2_p) \\
            \boldsymbol{\gamma}_{b_i} &\sim \mathcal{N}(\boldsymbol{0}, \Sigma)
        \end{split}
    \end{align}
    where $\boldsymbol{\gamma}_{b_i}$ is the vector of batter random intercept and slopes $(\gamma_{b_i}, \gamma^S_{b_i}, \gamma^X_{b_i}, \gamma^Z_{b_i})$ which follow a multivariate normal distribution centered around the zero vector $\boldsymbol{0}$ with covariance matrix $\Sigma$. This $\Sigma$ contains the relevant random effect variances for the batter-level terms, $(\sigma_b, \sigma_b^S, \sigma_b^X, \sigma_b^Z)$, as well as their respective correlations. This allows us to quantify variation between hitters and pitchers at the intercept level, as well as capture variation in batter approach and swing adaptation.

@@

    We fit both bat speed and swing length models in a Bayesian framework via the \texttt{brms} package in \texttt{R} \citep{brms}, which provides an interface for Bayesian modeling with Stan \citep{carpenter2017stan}. We use weakly informative priors for the parameters, with vague half-$t_3$ priors (i.e., a Student's $t$ distribution centered at zero with 3 degrees of freedom but truncated to positive values) for the standard deviation parameters \citep{gelman2006prior} and the LKJ prior for the correlations between batter-level effects \citep{lewandowski2009generating}. Our Bayesian approach naturally provides uncertainty quantification for the model parameters via their posterior distributions, which are estimated using MCMC sampling. For model fitting, we use four parallel chains with 6,000 (3,000 burn-in) and 4,000 (2,000 burn-in) samples for the swing length and bat speed models respectively. We observe evidence of convergence of the MCMC algorithm based on observing trace plots and $\hat{R}$ statistics close to 1 \citep{gelman1992inference, brooks1998general}, along with no indication of problematic effective sample sizes \citep{gelman2020bayesian}.

    For comparison, we also fit simpler Gaussian versions of the swing length and bat speed models. These models follow a specification similar to that of the SK models in Equation (\ref{eqn:intention-swing-length}), but with the shape parameter removed (i.e., the Guassian distribution is a SK with shape equal to 0). We use an 80/20 train/test split to compare the out-of-sample performance of the SK and Gaussian fits against each other. We compute the expected log pointwise predictive density (ELPD) on the 20\% test data, which involves computing the log-likelihood of each test observation for each posterior sample. In addition to providing us with standard errors about the difference in model performance, we choose ELPD as our evaluation metric because of our interest in comparing the skewness of the SK posterior predictive distributions relative to the symmetric Gaussian.

@@

      The out-of-sample ELPD results are shown in Table \ref{tab:intention-model-elpd} (SP: COULD WE REDUCE THE NUMBER OF DIGITS IN THIS TABLE?) for both swing length and bat speed models. For both settings, the SK model displayed an ELPD of around 3 standard errors better than the Gaussian model. The remaining results in this manuscript are based on the SK intention models.

      \begin{table}[H]
        \centering
        \begin{tabular}{l|r|r|r|}
          & $\Delta$ELPD  & SE & \# SEs \\
          \hline
          \input{../../tables/intention_elpd.tex}
        \end{tabular}
        \caption{\it Model comparison via expected log predictive density (ELPD). Shown are the difference in ELPD between the skew-normal (SK) and Gaussian models ($\Delta$ELPD), the standard error of ELPD difference (SE), and the number of standards errors the SK model is better than the Gaussian model.}
        \label{tab:intention-model-elpd}
      \end{table}

       Table \ref{tab:fixed-effects} summarizes the fixed effects of the bat speed and swing length intention models. We observe intuitive relationships for both models with regard to the relationships between pitch-level context and each observed swing measurement. As the number of balls increase ($\beta^B$), on average we observe faster and longer swings. The opposite behavior is observed as the number of strikes increases ($\beta^S$), with slower and shorter swings on average. Figure \ref{fig:approach} displays the joint distribution of the posterior mean for the batter random slopes for strikes ($\hat{B}^S + \hat{\gamma}_b^S$) in both intention models. Regardless of the individual batter effect we expect a decrease in bat speed and swing length per strike. A similar relationship is observed for pitches with higher vertical locations ($\beta^Z$). The two models diverge with regard to the horizontal location of pitches ($\beta^X$), as pitches that thrown further away from the batter result in longer but slower swings on average. With regard to the shape parameter of the SK distribution, the intercepts ($\alpha_0$) for both bat speed and swing length are negative, which indicate left skewed distributions.

      \begin{figure}[H]
        \centering
        \includegraphics[width = 0.6\textwidth]{../../figures/approach.pdf}
        \caption{\it Estimated batter random slopes for strikes in the intention model. Each point is a batter $b$, with the x-value representing $\hat\beta^S + \hat\gamma^S_b$ from the model (\ref{eqn:intention-swing-length}) and the y-value representing the same quantity from the analogous model for intended bat speed. This quantity is interpretable as the expected change in swing length (or bat speed) corresponding to a one-strike increase in the ball-strike count.}
        \label{fig:approach}
      \end{figure}

@@

      Next, Table \ref{tab:intention-variances} summarizes the posterior distributions of the random effects variance terms. For both bat speed and swing length intention models, we observe the highest variation is attributed to differences between batters ($\sigma_b$). Interestingly, the variation between batters in the shape random effects ($\tau_b$) is the second highest for swing length but is smaller than the pitch location random slopes ($\sigma_b^X$ and $\sigma_b^Z$) for the bat speed model. For each of the estimated terms, all of the 95\% credible intervals are strictly greater than zero demonstrating evidence of variability between batters at the intercept and slope-level. (SP: DOES OUR PRIOR PUT ANY WEIGHT ON ZERO, OR WAS THIS FACT INEVITABLE?)

@@

      Figure \ref{fig:intent-re} provides an overview of the batter-level $\mu_i$ random effects in the SK intention models. The variation we observe between batters aligns with intuition, as batters known for taking longer/faster swings (e.g., Oneil Cruz, Giancarlo Stanton, Javier Baez) display higher random intercepts ($\gamma_b$) in comparison with batters known for shorter/slower swings (e.g., Luis Arraez). We observe positive correlations between the random slopes of bat speed and swing length, providing information on the variation between batters with respect to their approach ($\gamma_b^S$) and adaptation ($\gamma_b^X$, $\gamma_b^Z$). For example, Anthony Rizzo's approach appears to be more conservative with the biggest decrease in bat speed and swing length for each additional strike relative to the average batter slope. We also observe batters that deviate from the consistent pattern between bat speed and swing length, such as Shohei Ohtani with regard to his horizontal pitch location random slope appearing to be higher than average for swing length despite a lower than average value for bat speed.

      \begin{figure}[H]
        \centering
        \includegraphics[width = \textwidth]{../../figures/intent_re.pdf}
        \caption{\it Joint distribution of the posterior means for the batter-level $\mu_i$ random effects in the SK intention models for both bat speed and swing length. Each batter is displayed by their name, and only batters with at least 25 squared-up swings are displayed.}
        \label{fig:intent-re}
      \end{figure}

       In contrast, Figure \ref{fig:alpha-re} displays the joint distribution of the posterior means for the batter random intercepts in modeling the shape parameter $\alpha$ of the SK distribution. Unlike the mean-level random effects, we do not observe correlation between the bat speed and swing length shape-level random intercepts. The variation captured by these random intercepts provides some understanding of the differences in the intended swing distribution for players. Luis Arraez displays above average random intercepts for both bat speed and swing length, indicating that his intention distributions are less skewed left than those of other batters.

@@

      The previous section demonstrates that there are differences between batters in how they modulate their swing length and bat speed as the number of strikes in the count increases. This section addresses the consequences of those differences between batters. As described in Section \ref{sec:methods-causal}, we estimate three separate regression models with instrumental variables for contact probability given swing (\ref{eqn:causal-contact}), fair ball probability given contact (\ref{eqn:causal-fair}) and expected xwOBAcon given fair ball (\ref{eqn:causal-hit}). The estimated effects of bat speed appraoch and swing length approach from these models are reported in Table \ref{tab:causal-model}.

      \begin{table}[H]
        \centering
        \begin{tabular}{l|r|r|r|}
          & Contact Model (\ref{eqn:causal-contact}) & Fair/Foul Model (\ref{eqn:causal-fair}) & xwOBAcon Model (\ref{eqn:causal-hit})\\
          \hline
          \input{../../tables/causal_model.tex}
        \end{tabular}
        \caption{\it Estimated regression coefficients and corresponding standard errors from the causal models (\ref{eqn:causal-contact})--(\ref{eqn:causal-hit}). Swing Length Approach is defined analogously as the batter's change in swing length per strike added to the count, as estimated by (\ref{eqn:intention-swing-length}). Bat Speed Approach is defined as the batter's change in bat speed per strike added to the count, as estimated by the analagous model for intended bat speed.}
        \label{tab:causal-model}
      \end{table}

@@

    In this work, we use newly available bat tracking data and examine the relationship between swing aggression and outcomes in baseball. We recognize that both bat speed and swing length are inherently outcomes of the swing, confounded by whether or not the batter identified the pitch. To address this, we first model the batter's bat speed and swing length intent via Bayesian hierarchical skew-normal models with evidence of more appropriate fits compared to simpler Gaussian models. Our results demonstrate the variation between batters in multiple aspects, including how they change their swing metrics as the number of strikes increases. We then condition on the batter's count-intended bat speed and swing length as instrumental variables to estimate the causal effect of these intended swing metrics on swing outcomes. The results are intuitive, indicating that batters who shorten and slow down their swing more per strike are expected to make contact more often in 1-strike and 2-strike counts relative to other batters. However, we observe a tradeoff as this type of conservative approach at the plate is associated with less power. We quantify this tradeoff in terms of run value, revealing on an interpretable scale that the highest-value approaches cut down on swing length without sacrificing bat speed.

@@

    While Section \ref{sec:results-value} illustrated the tradeoff involved with how batters vary their \textit{approach}, there are of course other sources of variation in bat speed and swing length. For example, in Section \ref{sec:methods-intention} we define \textit{adaptation} as the extent that a batter's swing tracking metrics co-vary with pitch location. As captured by $\sigma_b^X$ and $\sigma_b^Z$ in Table \ref{tab:intention-variances} and the posterior mean for the batter level random effects in Figure \ref{fig:intent-re}, we observe differences between batters in their intended \textit{adaptation} as the pitch location varies horizontally and vertically. To demonstrate the variation, Figure \ref{fig:adaptation} displays the expected swing length from the intention model by pitch location for two sample batters, Nico Hoerner and Juan Soto (assuming a 0-0 count). Hoerner displays a clear relationship with the vertical location with shorter intended swings on average for pitches that are located higher in the zone. In comparison, Soto displays a fairly consistent swing length throughout the zone with an uptick in length for low and away pitches, versus a shortening of the swing for high and inside pitches. Understanding why batters display such variation in their \textit{adaptation} is left for future work, and could be explored with swing path measurements describing the tilt and attack angle that were released during the 2025 MLB season \citep{petriello_4_2025}.

@@

      We limited our analysis of the run value of approach to a batter of average skill, but it is possible that the results in Section \ref{sec:results-value} could differ depending on the profile of the batter. For example, the value of a specific approach could be different for high-contact, low-power batters versus low-contact, high-power batters. Future work could estimate personalized effects of batter approach based on the batter's strengths and weaknesses. Additionally, the estimated run value was based on linear weights, which ignore the base-out situation. With a runner on third base on less than two outs, the value of putting the ball in play (relative to non-ball-in-play outcomes) increases, so presumably the tradeoff of increased contact for decreased power is more valuable in this situation. Future work could estimate this effect and evaluate the extent to which batters tailor their approach to the base-out situation.
